---
title: 'Using traffic'
author: "Koen Van den Berge, Hector Roux de Bézieux, Kelly Street"
bibliography: traffic.bib
date: "03/06/2021"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{traffic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
```

# Introduction

The `traffic` package was created to collect visualization functions related to trajectory inference. Asides from general purpose functions useful to any user, it also contains visualizations that can be used in a `slingshot`, `tradeSeq` or `condiments` workflow.

# Visualization of a trajectory

First, we will demonstrate functions to visualize a trajectory. Here, we'll work with a trajectory as estimated by Slingshot [@Street2018]. An example trajectory is provided along with the `traffic` package.

Below, we show how one can use the `lines` and `plot` functions for a `SlingshotDataSet` object. We also show how the `plotGeneCount` function can be used for a quick visualization of the trajectory.

```{r}
library(slingshot)
library(traffic)
data(crv, package = "traffic")
class(crv)
rd <- slingReducedDim(crv)
cl <- apply(slingClusterLabels(crv),1, function(x) which(x==1))

## Only visualize the trajectory
plot(crv)

## Visualize the trajectory on top of cells in reduced space
plot(rd, pch=16, col=cl+1, cex=2/3)
lines(crv, col="black", lwd=3)

## Visualizing trajectory and clusters using plotGeneCount
plotGeneCount(crv, clusters=cl)
```

### 3D visualization

The same functions (i.e., `plot` and `lines`) can also be used to visualize trajectories in 3D space using the `rgl` package. This can be done using the `plot3d` and `lines3d` functions, in similar vein as the 2D visualizations above.

# Visualization of gene expression along a trajectory

## Visualizing gene expression along with the trajectory in reduced space

The `plotGeneCount` also allows you to visualize the trajectory in reduced space, where
each cell is colored according to its expression of the gene as defined by the `gene` argument.

```{r}
data(counts, package="traffic")
plotGeneCount(crv, counts, gene = "Mpo")
```

## Visualizing gene expression along pseudotime

The `plotExpression` function may be used to visualize gene expression along pseudotime along with a quick smooth estimate of average gene expression. It uses a loess smoother to smooth the raw gene expression values as a function of pseudotime. Note that no between-cell normalization occurs to do the smoothing.

```{r}
plotExpression(counts=counts,
               sds=crv,
               gene="Car1")
```

If smoothers have been fitted using `tradeSeq`, then one can use the `plotSmoothers` function to plot the estimated smoothers. Note that here normalization does happen within `tradeSeq`, and the smoothers are estimated using a negative binomial count model.

```{r}
data(sce, package="traffic")

plotSmoothers(models = sce, counts = counts, gene = "Elane")
```



# Feedback and suggestions

Any comment, question, or feedback is welcome and can be posted as an issue on the [traffic repo](https://github.com/koenvandenberge/traffic).

# Session Info

```{r}
sessionInfo()
```

# References
